---
title: "Caret Package"
author: "Thomas Scheuchzer <thomas.scheuchzer@gmx.net>"
date: "Monday, June 08, 2015"
output: html_document
---

## The caret R package

http://topepo.github.io/caret/

The caret package (short for Classification And REgression Training) is a set of functions that attempt to streamline the process for creating predictive models. The package contains tools for:

- data splitting
- pre-processing
- feature selection
- model tuning using resampling
- variable importance estimation

as well as other functionality.

## Caret functionality

* Some preprocessing (cleaning)
  * preProcess
* Data splitting
  * createDataPartition
  * createResample
  * createTimeSlices
* Training/testing functions
  * train
  * predict
* Model comparison
  * confusionMatrix


## Why caret? 

<img class=center src=predicttable.png height=250>


## SPAM Example: Data splitting

```{r loadPackage}
library(caret); library(kernlab); data(spam)
inTrain <- createDataPartition(y=spam$type,
                              p=0.75, list=FALSE)
training <- spam[inTrain,]
testing <- spam[-inTrain,]
dim(training)
```

## SPAM Example: Fit a model

```{r training, dependson="loadPackage",cache=TRUE}
set.seed(32343)
modelFit <- train(type ~.,data=training, method="glm")
modelFit
```

## SPAM Example: Final model

```{r finalModel, dependson="training",cache=TRUE}
modelFit <- train(type ~.,data=training, method="glm")
modelFit$finalModel
```

## SPAM Example: Prediction

```{r predictions, dependson="training",cache=TRUE}
predictions <- predict(modelFit,newdata=testing)
head(predictions)
```

## SPAM Example: Confusion Matrix

```{r confusion, dependson="predictions",cache=TRUE}
confusionMatrix(predictions,testing$type)
```

# Data Slicing

## SPAM Example: Data splitting

```{r}
library(caret); library(kernlab); data(spam)
inTrain <- createDataPartition(y=spam$type,
                              p=0.75, list=FALSE)
training <- spam[inTrain,]
testing <- spam[-inTrain,]
dim(training)
```


## SPAM Example: K-fold

Used for cross validation.

```{r kfold,dependson="loadPackage"}
set.seed(32323)
folds <- createFolds(y=spam$type,k=10,
                             list=TRUE,returnTrain=TRUE)
sapply(folds,length)
folds[[1]][1:10]
```

## SPAM Example: Return test

```{r kfoldtest,dependson="loadPackage"}
set.seed(32323)
folds <- createFolds(y=spam$type,k=10,
                             list=TRUE,returnTrain=FALSE)
sapply(folds,length)
folds[[1]][1:10]
```

## SPAM Example: Resampling

```{r resample,dependson="loadPackage"}
set.seed(32323)
folds <- createResample(y=spam$type,times=10,
                             list=TRUE)
sapply(folds,length)
folds[[1]][1:10]
```

## SPAM Example: Time Slices

```{r time,dependson="loadPackage"}
set.seed(32323)
tme <- 1:1000
folds <- createTimeSlices(y=tme,initialWindow=20,
                          horizon=10)
names(folds)
folds$train[[1]]
folds$test[[1]]
```

# Train Options

```{r cache=TRUE}
library(caret); library(kernlab); data(spam)
inTrain <- createDataPartition(y=spam$type,
                              p=0.75, list=FALSE)
training <- spam[inTrain,]
testing <- spam[-inTrain,]
modelFit <- train(type ~.,data=training, method="glm")
```

## Train options

```{r ,dependson="loadPackage"}
args(train.default)
```

## Metric options

__Continous outcomes__:
  * _RMSE_ = Root mean squared error
  * _RSquared_ = $R^2$ from regression models

__Categorical outcomes__:
  * _Accuracy_ = Fraction correct
  * _Kappa_ = A measure of [concordance](http://en.wikipedia.org/wiki/Cohen%27s_kappa)
  
## trainControl

```{r , dependson="loadPackage",cache=TRUE}
args(trainControl)
```  

## trainControl resampling

* _method_
  * _boot_ = bootstrapping
  * _boot632_ = bootstrapping with adjustment
  * _cv_ = cross validation
  * _repeatedcv_ = repeated cross validation
  * _LOOCV_ = leave one out cross validation
* _number_
  * For boot/cross validation
  * Number of subsamples to take
* _repeats_
  * Number of times to repeate subsampling
  * If big this can _slow things down_

## Setting the seed

* It is often useful to set an overall seed
* You can also set a seed for each resample
* Seeding each resample is useful for parallel fits


## seed example

```{r , dependson="seedExample",cache=TRUE}
set.seed(1235)
modelFit2 <- train(type ~.,data=training, method="glm")
modelFit2
```


