---
title: "Preprocessing"
author: "Thomas Scheuchzer <thomas.scheuchzer@gmx.net>"
date: "Monday, June 08, 2015"
output:
  html_document:
    keep_md: true
---


## Why preprocess?

```{r loadPackage,cache=TRUE,fig.height=3.5,fig.width=3.5}
library(caret); library(kernlab); data(spam)
inTrain <- createDataPartition(y=spam$type,
                              p=0.75, list=FALSE)
training <- spam[inTrain,]
testing <- spam[-inTrain,]
hist(training$capitalAve,main="",xlab="ave. capital run length")
```


## Why preprocess?

```{r ,dependson="loadPackage",cache=TRUE,fig.height=3.5,fig.width=3.5}
mean(training$capitalAve)
sd(training$capitalAve)
```


## Standardizing

```{r ,dependson="loadPackage",cache=TRUE,fig.height=3.5,fig.width=3.5}
trainCapAve <- training$capitalAve
trainCapAveS <- (trainCapAve  - mean(trainCapAve))/sd(trainCapAve) 
mean(trainCapAveS)
sd(trainCapAveS)
```


## Standardizing - test set

Important! For standardization you must only use calcualted values from the
training set! `trainCapAve`
```{r ,dependson="loadPackage",cache=TRUE,fig.height=3.5,fig.width=3.5}
testCapAve <- testing$capitalAve
testCapAveS <- (testCapAve  - mean(trainCapAve))/sd(trainCapAve) 
mean(testCapAveS)
sd(testCapAveS)
```


## Standardizing - _preProcess_ function
`preProcess()` is from `caret`.

`center` and `scale` is the same as we did in `Standardizing`.
```{r preprocess,dependson="loadPackage",cache=TRUE,fig.height=3.5,fig.width=3.5}
preObj <- preProcess(training[,-58],method=c("center","scale"))
trainCapAveS <- predict(preObj,training[,-58])$capitalAve
mean(trainCapAveS)
sd(trainCapAveS)
```



## Standardizing - _preProcess_ function

Apply same preprocessing to the test set.
```{r ,dependson="preprocess",cache=TRUE,fig.height=3.5,fig.width=3.5}
testCapAveS <- predict(preObj,testing[,-58])$capitalAve
mean(testCapAveS)
sd(testCapAveS)
```


## Standardizing - _preProcess_ argument

```{r training, dependson="loadPackage",cache=TRUE}
set.seed(32343)
modelFit <- train(type ~.,data=training,
                  preProcess=c("center","scale"),method="glm")
modelFit
```



## Standardizing - Box-Cox transforms

```{r ,dependson="loadPackage",cache=TRUE,fig.height=3.5,fig.width=7}
preObj <- preProcess(training[,-58],method=c("BoxCox"))
trainCapAveS <- predict(preObj,training[,-58])$capitalAve
par(mfrow=c(1,2)); hist(trainCapAveS); qqnorm(trainCapAveS)
```


## Standardizing - Imputing data

```{r knn,dependson="loadPackage",cache=TRUE,fig.height=3.5,fig.width=7}
set.seed(13343)

# Make some values NA
training$capAve <- training$capitalAve
selectNA <- rbinom(dim(training)[1],size=1,prob=0.05)==1
training$capAve[selectNA] <- NA

# Impute and standardize
preObj <- preProcess(training[,-58],method="knnImpute")
capAve <- predict(preObj,training[,-58])$capAve

# Standardize true values
capAveTruth <- training$capitalAve
capAveTruth <- (capAveTruth-mean(capAveTruth))/sd(capAveTruth)
```



## Standardizing - Imputing data

```{r ,dependson="knn",cache=TRUE,fig.height=3.5,fig.width=7}
quantile(capAve - capAveTruth)
quantile((capAve - capAveTruth)[selectNA])
quantile((capAve - capAveTruth)[!selectNA])
```


## Notes and further reading

* Training and test must be processed in the same way
* Test transformations will likely be imperfect
  * Especially if the test/training sets collected at different times
* Careful when transforming factor variables!
* [preprocessing with caret](http://caret.r-forge.r-project.org/preprocess.html)

