---
output: html_document
---
# Raw and Processed Data



# Components of Tidy Data

## Raw Data

It's raw data when:

- Ran no software on the data
- Didn't manipulate any of the numbers in the data
- You didn't remove any data from the data set
- You didn't summarize the data in any way 

## Tidy Data Set

- Each variable is in a separat column
- Each different observation is in a separate row
- One table for each "kind" of varable (one table per source)
- Primary key if you have multiple tables


## Code Book

The code book contains the meta-data of your tidy data. It describes each column/variable.

- There should be a section called `Study design` that has a through description of how you collected data
- There must be a sectionc alled `Code book` that describes each variable and its units


## Instruction List

- Ideally a script
- The input for the script is the raw data
- The output is processed, tidy data
- There are no parameters to the script

or describe manual scripts step by step.



## Downloading Files

```r
download.file(url,destfile, method	)
```

```{r}
if(!file.exists("data")) {
	dir.create("data")
}
url <- "https://data.baltimorecity.gov/api/views/dz54-2aru/rows.csv?accessType=DOWNLOAD"
download.file(url, destfile="./data/cameras.csv", method="curl")
list.files("./data")
```



## Reading Local Files

```{r}
data <- read.csv("./data/cameras.csv")
head(data)
```



## Reading Excel Files

```bash
sudo aptitude install r-cran-rjava
```
```r
install.packages('xlsx')
```

```{r}
url <- "https://data.baltimorecity.gov/api/views/dz54-2aru/rows.xlsx?accessType=DOWNLOAD"
download.file(url, destfile="./data/cameras.xlsx", method="curl")
list.files("./data")

library("xlsx")
data <- read.xlsx("./data/cameras.xlsx", sheetIndex = 1, header = TRUE)
head(data)
```

Also see package `XLConnect` for an alternative.



## Reading XML Files

```r
install.packages("XML")
```

```{r}
library(XML)
url <- "http://www.w3schools.com/xml/simple.xml"
doc <- xmlTreeParse(url, useInternalNodes = TRUE)
rootNode <- xmlRoot(doc)    
xmlName(rootNode)
names(rootNode)
rootNode[[1]]
rootNode[[1]][[1]]
```

Apply a function to xml elements

```{r}
xmlSApply(rootNode, xmlValue)
```


### X-Path

```{r}
xpathSApply(rootNode, "//price", xmlValue)
```


## Reading HTML Files

```{r}
library(XML)
url <- "http://espn.go.com/nfl/team/_/name/bal/baltimore-ravens"
doc <- htmlTreeParse(url, useInternalNodes = TRUE)
xpathSApply(doc, "//li[@class='team-name']", xmlValue)
```



## Reading JSON Files

```{r}
library(jsonlite)
jsonData <- fromJSON("https://api.github.com/users/jtleek/repos")
names(jsonData)
```


## Writing JSON Files

```{r}
myJson <- toJSON(iris[1:2,], pretty=TRUE)
cat(myJson)
```



## data.table

### data frame
```{r}
library('data.table')
df <- data.frame(x=rnorm(9), y=rep(c("a", "b", "c"), each=3), z=rnorm(9))
head(df, 3)
```

### table

```{r}
library('data.table')
dt <- data.table(x=rnorm(9), y=rep(c("a", "b", "c"), each=3), z=rnorm(9))
head(dt, 3)
```

See information about all data tables in memory
```{r}
tables()
```

Calculate values for variables with expressions
```{r}
dt[, list(mean(x), sum(z))]

dt[, table(y)]
dt[, .N, by=y]
```

Add new columns. **Doesn't copy the values into a new table. Changes the current table!**
```{r}
dt[, w:=z^2]
dt
```

Multiple operations
```{r}
dt[, w:={tmp <- (x+z); log2(tmp+5)}]
dt
```

Add booleans depending on a variable
```{r}
dt[, a:=x>0]
dt
```


Group by a variable. Adds the mean of x+w grouped by a. So all rows with the same a will have the same b
```{r}
dt[, b:=mean(x+w), by=a]
dt
```

Keys. Allows selection without expression for the column
```{r}
setkey(dt, y)
dt['b']
```

Join
```{r}
dt1 <- data.table(x=c('a','a','b', 'dt1'), y=1:4)
dt2 <- data.table(x=c('a', 'b', 'dt2'), z=5:7)
setkey(dt1, x)
setkey(dt2, x)
merge(dt1, dt2)
```

fast table reading
```r
fread(file)
```

