---
title: "Colors"
output: html_document
---

The default colors for plotting are not very nice. There are some color palettes that can be loaded.

- rainbow
- heat.colors
- terrain.colors
- topo.colors
- cm.colors

```{r}
filled.contour(volcano, color.palette = terrain.colors, asp = 1)
filled.contour(volcano, col = terrain.colors(5), asp = 1)
filled.contour(volcano, color.palette =topo.colors , asp = 1)
```


# grDevices Package

```{r}
library(grDevices)
```

- colors(): returns the build-in color names
- colorRamp(): returns a function that interpolates a set of given colors to create new color palettes.
- colorRampPalette()
- gray(): create a vector of colors from a vector of gray levels. Same as `colorRampPalette(c("black","white"))` but as hex.

```{r}
pal <- colorRamp(c("red", "blue"))
pal(0)
pal(0.5)
pal(1)
```

To get hex colors call the `colorRampPalette` function. This function takes integers not values between 0 and 1. It generates a vector containing the amount of color values between the given two colors.

```{r}
pal <- colorRampPalette(c("black","white"))
pal(1)
pal(2)
pal(3)
```

# RColorBrewer Package

Usefule color palettes.

3 types of palettes:

- sequential
- diverging
- qualitative

Palette information can be use in conjunction with the `colorRamp` and `colorRampPalette` functions.

```{r}
library(RColorBrewer)
cols <- brewer.pal(3, "BuGn")
cols
pal <- colorRampPalette(cols)
image(volcano, col=pal(20))
```

Find the name of the palette on the help page!


```{r}
cols <- brewer.pal(9, "Greens")
x <- rnorm(10000)
y <- rnorm(10000)
smoothScatter(x,y, colramp = colorRampPalette(cols))
```

# other functions

## rgb

returns the hex value for a color

```{r}
rgb(1, 0.5, 0)
rgb(1, 0.5, 0, alpha = 1)
```

## transparency

```{r}
x <- rnorm(5000)
y <- rnorm(5000)
plot(x,y, pch=19)
plot(x,y, col = rgb(0,0,0,0.2), pch=19)
```