---
title: "EDA Case Study"
date: "23.02.2015"
output: html_document
---

Data: http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones

```{r}
load("samsungData.rda")
names(samsungData)[1:12]
table(samsungData$activity)
```

transform activity into a factor
```{r}
par(mfrow = c(1,2), mar = c(5, 4, 1, 1))
samsungData <- transform(samsungData, activity = factor(activity))
```

select subject 1 and display first two variables
```{r}
sub1 <- subset(samsungData, subject == 1)
plot(sub1[, 1], col = sub1$activity, ylab = names(sub1)[1])
plot(sub1[, 2], col = sub1$activity, ylab = names(sub1)[2])
legend("bottomright", legend = unique(sub1$activity), col = unique(sub1$activity), pch = 1)
```

clustering based on the average acceleration
```{r}
source("myplclust.R")
distanceMatrix <- dist(sub1[, 1:3])
hclustering <- hclust(distanceMatrix)
par(mfrow = c(1,1))
myplclust(hclustering, lab.col = unclass(sub1$activity))
```
looks messy, no real information with just this data

Plotting the max acceleration for the first subject
```{r}
par(mfrow = c(1, 2))
plot(sub1[, 10], pch = 19, col = sub1$activity, ylab = names(sub1)[10])
plot(sub1[, 11], pch = 19, col = sub1$activity, ylab = names(sub1)[11])
legend("bottomright", legend = unique(sub1$activity), col = unique(sub1$activity), pch = 19)
```

clustering based on the max acceleration
```{r}
distanceMatrix <- dist(sub1[, 10:12])
hclustering <- hclust(distanceMatrix)
par(mfrow = c(1,1))
myplclust(hclustering, lab.col = unclass(sub1$activity))
```
separates moving from non moving but not much else

singular value decomposition
```{r}
par(mfrow = c(1, 2))
svd1 <- svd(scale(sub1[, -c(562, 563)])) # remove last two rows, no real data
# first left singular vector
plot(svd1$u[, 1], col = sub1$activity, pch = 19)
# first left singular vector
plot(svd1$u[, 2], col = sub1$activity, pch = 19)
legend("bottomright", legend = unique(sub1$activity), col = unique(sub1$activity), pch = 19)
```
The first singular vector seems to separate moving from non moving.

Find maximum contributor
```{r}
par(mfrow = c(1,1))
plot(svd1$v[, 2], pch = 19)
maxContrib <- which.max(svd1$v[, 2])
points(x = maxContrib, y = svd1$v[maxContrib,2], pch = 19, col = "red")
```
Which feature is contributing the most variance between the different observations.

New cluster with maximum contributor
```{r}
maxContrib <- which.max(svd1$v[, 2])
distanceMatrix <- dist(sub1[, c(10:12, maxContrib)])
hclustering <- hclust(distanceMatrix)
myplclust(hclustering, lab.col = unclass(sub1$activity))
legend("topright", legend = unique(sub1$activity), col = unique(sub1$activity), pch = 19)
```
move activities better separated. Didn't help much to separate the non moving activities.

```{r}
names(samsungData)[maxContrib]
```

K-means clustering (nstart=1, first try)
```{r}
kClust <- kmeans(sub1[, -c(562, 563)], centers = 6)
table(kClust$cluster, sub1$activity)

kClust <- kmeans(sub1[, -c(562, 563)], centers = 6, nstart = 1)
table(kClust$cluster, sub1$activity)
#kmeans chooses random starting points, so we different results
kClust <- kmeans(sub1[, -c(562, 563)], centers = 6, nstart = 100)
table(kClust$cluster, sub1$activity)
kClust <- kmeans(sub1[, -c(562, 563)], centers = 6, nstart = 100)
table(kClust$cluster, sub1$activity)
```

Cluster 1 Variable Centers (Laying)
```{r}
set.seed(50)
kClust <- kmeans(sub1[, -c(562, 563)], centers = 6, nstart = 1)
plot(kClust$centers[1, 1:10], pch = 19, ylab = "Cluster Center", xlab = "")
```

Cluster 4 Variable Centers (Walking)
```{r}
set.seed(50)
kClust <- kmeans(sub1[, -c(562, 563)], centers = 6, nstart = 1)
plot(kClust$centers[4, 1:10], pch = 19, ylab = "Cluster Center", xlab = "")
```

Looking at cluster centers might give you a hint which features drive the location of the center. 