---
title: "Exploratory Graphs"
output: html_document
---

Graphs that you make more or less for yourself.

# Why graphs?

- To understand data properties (exploratory graphs)
- To find patterns in data (exploratory graphs)
- To suggest modelling strategies (exploratory graphs)
- To debug analyses (exploratory graphs)
- To communicate results


# Characteristics of exploratory graphs

- made quickly
- a large number are made
- the goal is for personal understanding
- axes/legends are generally cleaned up later
- color/size are primairly used for information

# Simple summaries of data

## One dimension

- Five-number summary
    - min, 1st qu., median, 3rd qu., max, (and mean)
    - `summary(data)`
- Boxplots
- Histograms
- Density plot
- Barplot

### Histogram with rug

```{r}
data <- rnorm(100)
hist(data, col='green')
rug(data)
```

### Histogram with overlaying features

```{r}
hist(data, col='green')
abline(v = 2, lwd=2)
abline(v = median(data), lwd=2, col="magenta")
```

### Boxplot

```{r}
boxplot(data, col='blue')
# add a line at 1.5
abline(h = 1.5)
```



## Two dimension

- Multiple/overlayed 1-D plots (Lattice/ggplot2)
- Scatterplots
- Smooth scatterplots

### Multiple histograms

```{r}
data1 <- rnorm(100)
data2 <- rnorm(100)
par(mfrow = c(2,1), mar = c(4,4,2,1))
hist(data1, col='green')
hist(data2, col='green')
```


### Scatterplots with more dimensions

```{r}
par(mfrow = c(1,1))
df1 <- data.frame(value=data1, region=sample(rep(1:2, 50)))
# add an additional dimension by coloring the region column
with(df1, plot(value, col = region))
    ```
## >2 dimension

- Overlayed/multiple 2-D plots; coplots
- Use color, size to add dimension
- Spinning plots
- Actual 3-D plots (not that useful)



# Resources

- http://rgraphgallery.blogspot.ch/
- http://www.r-bloggers.com/