---
title: "Plotting Systems"
output: html_document
---

There are three plotting systems. 

- The Base Plotting System
- The Lattice System
- The ggplot2 System

Don't mix them!



# The Base Plotting System

- "Artist's palette" model
- start with a blank canvas and build up from there
- start with `plot()` or similar
- use annotation functions to add/modify
  - text
  - lines
  - points
  - axis
  
Good for exploration. But:

- can't go back
- just a series of R commands 
    - keep track of them when you explore
- difficult to translate to others once it has been plotted 
    - no graphical language
    

```{r}
library(datasets)
data(cars)
with(cars, plot(speed, dist))
```


The base plotting system is implemented in two packages:

- `graphics`: functions like `plot`, `hist`, `boxplot`
- `grDevices`: graphics devices like X11, PDF, PNG...

## Configuration

The graphing system can be configured through the `par()` function (global) or as function argument of the plot function.

Important plot parameters:

- pch: the plotting symbol
- lty: line type
- lwd: line width
- col: the plotting color
    - number, string or hex
    - `colors()` gives you a vector of colors by name
- xlab: x-axis label
- ylab: y-axis label

Important `par` parameters:

- las: orientation of the axis labels on the plot
- bg: background color
- mar: margin size
    - bottom, left, top, right
- oma: outer margin size
- mfrow: number of plots per row
- mfcol: number of plots per column

Get the defaults:

```{r}
par("bg")
```


## Histogram

```{r}
library(datasets)
hist(airquality$Ozon)
```

## Scatterplot

```{r}
library(datasets)
with(airquality, plot(Wind, Ozone))
```

## Boxplot

```{r}
library(datasets)
airquality <- transform(airquality, Month = factor(Month))
boxplot(Ozone ~ Month, airquality, xlab = "Month", ylab = "Ozone (ppb)")
```


## Annotation functions

- `lines`
- `legend`
- `points`
- `text`: add text to x,y coordinates
- `title`: x,y labels, title, subtitle, outer margin
- `mtext`: add text to the margins of the plot 
- `axis`: ticks and labels
- `abline`

## Base plot with annotation

We start with an empty plot (`type="n"`) as we paint in two steps without overpainting.

```{r}
with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in New York City", type="n"))
with(subset(airquality, Month == 5), points(Wind, Ozone, col="blue"))
with(subset(airquality, Month != 5), points(Wind, Ozone, col="red"))
legend("topright", pch = 1, col = c("blue", "red"), legend = c("May", "Other months"))
```

## Base plot with regression line

```{r}
with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in New York City", pch = 20))
model <- lm(Ozone ~ Wind, airquality)
abline(model, lwd = 2)
```


## Multiple base plots

```{r}
par(mfrow = c(1, 2))
with(airquality, {
    plot(Wind, Ozone, main = "Ozone and Wind")
    plot(Solar.R, Ozone, main = "Ozone and Solar Radiation")
})
```


## Plot to a graphics device

### PDF

```{r}
library(datasets)
pdf(file = "test.pdf")
par(mfrow = c(1,1))
with(faithful, plot(eruptions, waiting))
title(main = "Old Faithful Geyser data")
dev.off()
```

### Multiple graphics devices

- You can open multiple graphics devices
- Every open graphics device is assigned an integer >= 2

```{r}
# Current device
dev.cur()
# Change the current device
dev.set(2)
```

### Copy plots

Copy from screen to file. Might not look exactly the same!

```{r}
library(datasets)
with(faithful, plot(eruptions, waiting))
title(main = "Old Faithful Geyser data")
dev.copy(png, file = "test.png")
dev.off()
```

## R integrated demo

```r
example(points)
```

# The Lattice System

```r
library("lattice")
```
- Single function call for creation
    - `xyplot`, `bwplot`, `histogram`, `stripplot`, `dotplot`, `splom`, `levelpolot`, `contourplot`
- most useful for conditioning types of plots
    - looking at how y changes with x across levels of z
- margins, spacing... set automatically because entire plot is specified at once
- good for putting many plots on a screen

Drawbacks:

- sometimes ackward to specify an entire plot in one function call
- annotation in plot is not especially intuitive
- you can't modify the plot after creation

```{r}
library(lattice)

state <- data.frame(state.x77, region = state.region)
xyplot(Life.Exp ~ Income | region, data = state, layout = c(4,1))
```


The base plotting system is implemented in two packages:

- `lattice`: functions like `xyplot`, `bwplot`, `levelplor`
- `grid`: the actual implementation of the graphing system on which lattice is based on. Lattice is independend from the base graphing system


## general usage

```r
xyplot( y ~ x | f * g, data)
```

- `y ~ x` the y- and x-axis variable
- `f * g` conditional variables (optional)
    - `*` indicates that there is an interaction between two variables
    
lattice functions return objects from type `trellis`. As we usually don't store the variable the return objects
gets `print()`-ed. This is when the plot actually is sent to the graphics device.

## xyplot()

```{r}
library(lattice)
library(datasets)
xyplot(Ozone ~ Wind, data = airquality)
```

Now month as factor

```{r}
airquality <- transform(airquality, Month = factor(Month))
xyplot(Ozone ~ Wind | Month, data = airquality, layout = c(5,1))
```


## Panel function

```{r}
set.seed(10)
x <- rnorm(100)
f <- rep(0:1, each = 50)
y <- x + f - f * x + rnorm(100, sd = 0.5)
f <- factor(f, labels = c("Group1", "Group2"))
xyplot(y ~ x | f, layout = c(2, 1))
```

Now with panel function that is called for rendering.

```{r}
xyplot(y ~ x | f, panel = function(x,y, ...) {
    panel.xyplot(x, y, ...)
    panel.abline(h = median(y), lty = 2)
})
```

```{r}
xyplot(y ~ x | f, panel = function(x,y, ...) {
    panel.xyplot(x, y, ...)
    panel.lmline(x, y, col = 2)
})
```








# The ggplot2 System

```r
install.packages("ggplot2")
```

Implementation of the "Grammar of Graphics" book.

- needs data frames
- splits the differences between base and lattice in a number of ways
- automatically deals with spacing, text, titles but also allows you to annotate by adding to a plot
- superficial similarity to lattice but generally easier/more intuitive to use
- default mode makes many choices for you
    - but you can still customize
    
Best practices:

- use factor variables. ggplot uses the levels for grouping and legends.
- read the documentation at: http://docs.ggplot2.org/current/ 

## Scatterplots

```{r}
library(ggplot2)
data(mpg)
qplot(displ, hwy, data = mpg)
qplot(displ, hwy, data = mpg, color = drv)
```

Select what to display by defining the `geom` attribute. Here we want to see the data points (`point`) and the trend (`smooth`). The `point` is the default geom if the attribute isn't specified.

```{r}
qplot(displ, hwy, data = mpg, geom = c("point", "smooth"))
```

## Histogram

```{r}
qplot(hwy, data = mpg, fill = drv)
```

## Facets (Panels)

Syntax: `facets = rows ~ columns`

If we just have one variable simply write a dot `.`.

```{r}
qplot(displ, hwy, data = mpg, facets = . ~ drv)
qplot(hwy, data = mpg, facets = drv ~ ., binwidth = 2)
```

## ggplot2 base function

A ggplot graphic consists out of several layers containing:

- a data frame
- aesthetic mappings: how data are mapped to color, size
- geoms: points, lines, shapes
- facets: for conditional plots
- stats: statistical transformations like binning, quantiles, smoothing
- scales: what scale a aesthetic map uses (example: male=red, female=blue)
- coordinate system

```{r}
g <- ggplot(mpg, aes(displ, hwy))
# can't be plotted as we don't have any geoms defined, yet.
p <- g + geom_point()
print(p)
p + geom_smooth(method = "lm") + theme_bw(base_family = "serif")
g + geom_point() + facet_grid(manufacturer~drv) + geom_smooth(method = "lm")
```

### Aesthetics

```{r}
g <- ggplot(mpg, aes(displ, hwy))
g + geom_point(color = "steelblue", size = 4, alpha = 1/2)
g + geom_point(aes(color = drv), size = 8, alpha = 1/4) + labs(title = "The Title")
g + geom_point(aes(color = drv), size = 8, alpha = 1/4) + geom_smooth(size = 2, linetype = 2, method = "lm", se = FALSE)
```

### Axis limits

With the base library. We don't care about the outlier. The interesting data is between -3 and 3.
```{r}
testdata <- data.frame(x = 1:100, y = rnorm(100))
testdata[50,2] <- 100 #Outlier!
plot(testdata$x, testdata$y, type="l", ylim = c(-3, 3))
```

With ggplot2. By default, all values are displayed.
```{r}
g <- ggplot(testdata, aes(x = x, y = y))
g + geom_line()
```

The `ylim()` function removes data that is not inside the boundaries. This might not be what you really want.
```{r}
g + geom_line() + ylim(-3, 3)
```

To get the same result as with the base plot we have to limit the coordinate system instead.
```{r}
g + geom_line() + coord_cartesian(ylim = c(-3,3))
```